\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-15 create-FSHybridPLS.Rmd
problem instance as matrix tuples

compare xi_hat_lin, xi_hat_lin_new

>>>>

### helper: get_xi_hat_linear
```{r}
get_xi_hat_eigen <- function(W, y){
  M <- W$n_basis_list[1]
  n <- W$n_sample
  K <- W$n_functional
  p <- W$n_scalar

  Theta_t <- W$functional_list[[1]]$coefs
  B <- W$gram_list[[1]]
  mat <- B %*% Theta_t %*% y
  for (i in 1:K){  ## tPhi_tC : C multiplied by Phi. Scalable to arbitrary K
    Theta_t <- W$functional_list[[i]]$coefs
    B <- W$gram_list[[i]]
    B_theta_y <- B %*% Theta_t %*% y
    mat <- rbind(mat, B_theta_y )
  }
  Zty<- t(W$Z) %*% y
  mat <- rbind(mat, Zty)
  mat <- mat %*% t(mat)
  geigen(mat, B, symmetric, only.values=FALSE)
  return(geigen)
}
```

### 2025-07-15 manuscript/main.tex
problem instance as matrix tuples

	\begin{remark}
		While it is possible to use a different basis for each functional predictor, we adopt a common basis system for all predictors for notational simplicity.
	\end{remark}

	Additionally, it serves as a regularization mechanism, borrowing strength across components and helping to prevent overfitting.

@@

	\begin{proposition}\label{proposition:generalized_rayleigh_quotient}
At the $l$-th iteration of the PLS algorithm, let $
( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top )^\top
$ denote the coefficients 
of $	\hat{\xi}$, 
defined in \eqref{def:maximizer_squared_empirical_cov},
which is
the projected unit-norm direction that maximizes the squared empirical covariance. These coefficients can be obtained by solving the following generalized eigenvalue problem
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$
where
$B^\ast := \operatorname{blkdiag}(B, \ldots, B, I_p)  \in  \mathbb{R}^{(MK+p) \times (MK+p)}$,
 $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$,
$B \in \mathbb{R}^{M \times M}$ is the gram matrix  of the basis functions  with entries
$
B_{m, m'} := \langle b_m, b_{m'} \rangle = \int_0^1  b_m(t)  b_{m'}(t) dt
$
for $m, m' = 1, \ldots, M$.
\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:generalized_rayleigh_quotient_linear}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:generalized_rayleigh_quotient} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{proposition:generalized_rayleigh_quotient_linear} are provided in Appendix~\ref{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

>>>>

The observations in the hybrid PLS problem can be compactly represented by the tuple
\begin{equation}\label{def:problem_instance}
	(B^\ast, \Theta, \mathbf{y}),
\end{equation}
defined as follows.
Let $B \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions, with entries  
\[
B_{m, m'} := \langle b_m, b_{m'} \rangle = \int_0^1 b_m(t)\, b_{m'}(t)\, dt,
\]
for $m, m' = 1, \ldots, M$.
Then define $B^\ast := \operatorname{blkdiag}(B, \ldots, B, I_p) \in \mathbb{R}^{(MK + p) \times (MK + p)}$,  
$\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$,  
and $\mathbf{y} := (y_1, \ldots, y_n)^\top$.
	\begin{remark}
While it is possible to use a different basis for each functional predictor, we adopt a common basis for all predictors to simplify the notation. However, the definition of $B^\ast$ remains general enough to accommodate different bases across the $K$ predictors.
\end{remark}

	%Additionally, it serves as a regularization mechanism, borrowing strength across components and helping to prevent overfitting.

@@

	\begin{proposition}\label{proposition:generalized_rayleigh_quotient}
At the $l$-th iteration of the PLS algorithm, let $
( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top )^\top
$ denote the coefficients 
of $	\hat{\xi}$, 
defined in \eqref{def:maximizer_squared_empirical_cov},
which is
the projected unit-norm direction that maximizes the squared empirical covariance.

 These coefficients can be obtained by solving the following generalized eigenvalue problem
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$

\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:generalized_rayleigh_quotient_linear}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:generalized_rayleigh_quotient} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{proposition:generalized_rayleigh_quotient_linear} are provided in Appendix~\ref{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.
